Species motif participation provides unique information about species risk of extinction

### 2022-06-19 manuscript/response_PRSB_round2.tex

        In addition, we believe that the Reviewer has over-estimated our reliance on the BioRxiv pre-print currently under review. Our previous draft had only three references to this pre-print: when describing the generation of simulated networks, when noting that degree and motif participation are not independent (which we also demonstrate \emph{de novo} in the present manuscript), and when briefly comparing the dynamical and Bayesian results in the discussion.
        We have now removed two of these references.
        We present the full network-generation methods in \emph{Appendix 1} and, to avoid confusion by referring to results presented in the preprint, have softened our language to say that degree and motif participation \emph{may} be correlated. 
        The latter is not a major point in our manuscript; we mention it only to emphasize that controlling for the total number of motifs in which a species appears does not necessarily control for degree. 
        Thus, we see some overlap in the associations of persistence with degree and with normalised motifs.

        We have retained the reference to the pre-print in the discussion because this brief comparison of results using dynamical and Bayesian simulations appears to be exactly what the Reviewer has asked for.
        As the Reviewer points out, different approaches to simulating disturbance can have different results, and we explicitly mention the main differences in the Discussion. 
        As the pre-print is freely available to any who wish to critically examine it (even if the peer-reviewed version is delayed), we believe that this is a good compromise between maintaining conciseness of the present manuscript and comparing different simulation approaches.
        If, all going well, the preprint is accepted for publication by the time the present manuscript is accepted, we will of course update the reference to the peer-reviewed version of the text.

### 2022-06-19 manuscript/reorder_main_text.tex
removing dead comments

    Species-level properties known to affect species' responses to disturbances include trophic level (a measure of the path length between a consumer and a basal resource) and in-degree (number of prey items of a predator). 
    These can also be referred to as local properties of the network.
    Species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally at higher risks of secondary extinction \citep{binzer2011susceptibility, Eklof2006}, and disturbances to plant populations cause particularly many secondary extinctions \citep{Curtsdotter2011}. %including secondary extinction due to reduced plant populations.
    However, in-degree and trophic level are both relatively coarse descriptions of a species' role in its food web~\citep{Cirtwill2018FoodWebs}. 
    A more detailed perspective is desirable since, as well as the properties of the consumer itself, a consumer's risk of secondary extinction after a disturbance to plants depends on the extinction risk of its prey.
    For example, a generalist consuming many specialist (and therefore high-risk) prey may be at greater risk of extinction than a specialist which consumes a highly generalist resource that is less likely to go extinct (Fig.~\ref{fig:concept}).

@@

%\subsection*{Here we test...}
    Here, we are interested in whether a consumer's extinction risk of going secondarily extinct depends on 1) the motifs in which the species participates, 2) the motif profile of the full food web and 3) the severity of the disturbance to the basal level. %We address this in scenarios with various levels of disturbance targeting the basal level.
    In addition, we want to understand how the information provided by motifs is related to other global and local food web properties in order to better synthesize information provided by different scales of network structure.
    We use synthetic networks and model secondary extinctions using Bayesian networks \citep{Eklof2013, Haussler2020}. 
    As Bayesian networks are computationally efficient (see Methods), we are able to simulate a wide range of disturbance intensities and test whether meso-scale network structure also interacts with severity of the disturbance.
    We focus our analyse on the four network motifs that are over-represented in empirical networks and additionally have been shown to contribute to network stability: the three-species chain, omnivory, apparent competition, and direct competition \citep{Stouffer2007, Borrelli2015a, giling2019plant}.

@@

    \subsection*{Species persistence and motif participation} 
    % Coefficients changed slightly with network-level random effect, but conclusions similar
       The relationship between frequency of participation in a motif and a species' probability of persistence depended on the level of disturbance. 
       When disturbance ($\pi$) was low, species' persistence increased with increasing proportions of the omnivory and direct competition motifs. 
       Conversely, persistence decreased with an increasing proportion of the apparent competition motif. 
       Persistence was not significantly related to the proportion of the three-species chain motif (Figure~\ref{fig:prop_lmer_all}; upper lines in each panel).

@@

    \subsection*{Motif participation and other network properties}
        % Updated with network-level random effect. No visible change.
        Species' motif participation varied depending on the global properties of the network: size, connectance, and their interaction (Fig. S1, Table S2; \emph{Appendix S5}).
        Species appeared most often in the apparent competition motif in all types of networks, but the proportion of apparent competition decreased with increasing network size and connectance.
        The proportion of the omnivory motif in a species' participation role increased sharply with increasing connectance, while the proportion of the direct competition motif increased slightly with increasing network size.

        % Updated with network-level random effect. No visible change.    
        Species' motif participation also varied with their in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}).
        Species with higher in-degree tended to have higher frequencies of participation in the omnivory motif. Proportions of the other three motifs tended to decline with increasing degree.
        Species with higher trophic levels tended to have higher proportions of apparent competition and three-species chains. 
        % Updated with network-level random effect. No visible change.   
        A species' persistence varies with both species in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}; \emph{Appendix S6}). 
        Increased trophic level is generally associated with decreasing species persistence, although the rate of decline is slower at high disturbance (Table S3, \emph{Appendix S6}).
        In-degree shows an even stronger interaction with disturbance; having more prey species is associated with higher persistence without disturbance, but lower persistence at high levels of disturbance.

@@

            %Update not needed
            When fitting regressions of persistence against motif participation for each network separately, networks with high connectance generally showed more consistent relationships between persistence and motif participation, i.e., sharper peaks of the density distributions (Fig.~\ref{fig:density_prop_C}).
            However, this trend varied at different levels of disturbance, especially for the direct competition and omnivory motifs.
            At low and medium connectances there was little change in the distribution of slopes for direct competition across all levels of disturbance. 
            At high connectances, however, the proportion of positive slopes was both lower overall and decreased with increasing disturbance. 
            For the omnivory motif these trends were reversed. 
            Moreover, the omnivory motif has a particularly broad distribution of slopes (i.e., an inconsistent relationship to persistence) when network connectance was low. 
            Network size had a much smaller effect than connectance (Fig. S4; \emph{Appendix S7}).

    \subsection*{Network mean persistence and motif profiles}
        % Updated to use means instead of all species
        The overall motif profile of a network was related to the average persistence of consumers ($F_{1,2998}$=645, $p$\textless0.001), although this result must be considered tentative due to non-homogeneous distribution of roles (\emph{Appendix S8}). 
        Average persistence was higher in networks with higher proportions of the three-species chain, direct competition, and apparent competition motifs (Table S4, \emph{Appendix S8}). 
        Conversely, average persistence was lower in networks with higher proportions of the omnivory motif. 
        These trends were consistent across levels of disturbance (Fig.~\ref{fig:motif_profile_persistence}, Table S4), \emph{Appendix S8}).% Network persistence varies between networks with different dispersion of motif profiles ($F_{134,2865}$=2.19$\times10^{24}$, $p$=\textless0.001). For example, networks with more variable motif profiles have higher persistence ($\beta$=0.183, $p$\textless0.001).

@@

% Appendices SA (persistence vs. S,C), SH-I (analyses with motif positions), SJ-K (repeated analyses with number of basal resources instead of disturbance) not referred to in text. Keeping SH-K in commented-out form in case reviewers ask. We should refer to SA somewhere in Discussion (will be Appendix 10).

>>>>

    Species-level properties known to affect species' responses to disturbances include trophic level (a measure of the path length between a consumer and a basal resource) and in-degree (number of prey items of a predator). 
    These can also be referred to as local properties of the network.
    Species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally at higher risks of secondary extinction \citep{binzer2011susceptibility, Eklof2006}, and disturbances to plant populations cause particularly many secondary extinctions \citep{Curtsdotter2011}. 
    However, in-degree and trophic level are both relatively coarse descriptions of a species' role in its food web~\citep{Cirtwill2018FoodWebs}. 
    A more detailed perspective is desirable since, as well as the properties of the consumer itself, a consumer's risk of secondary extinction after a disturbance to plants depends on the extinction risk of its prey.
    For example, a generalist consuming many specialist (and therefore high-risk) prey may be at greater risk of extinction than a specialist which consumes a highly generalist resource that is less likely to go extinct (Fig.~\ref{fig:concept}).

@@

%\subsection*{Here we test...}
    Here, we are interested in whether a consumer's extinction risk of going secondarily extinct depends on 1) the motifs in which the species participates, 2) the motif profile of the full food web and 3) the severity of the disturbance to the basal level. 
    In addition, we want to understand how the information provided by motifs is related to other global and local food web properties in order to better synthesize information provided by different scales of network structure.
    We use synthetic networks and model secondary extinctions using Bayesian networks \citep{Eklof2013, Haussler2020}. 
    As Bayesian networks are computationally efficient (see Methods), we are able to simulate a wide range of disturbance intensities and test whether meso-scale network structure also interacts with severity of the disturbance.
    We focus our analyse on the four network motifs that are over-represented in empirical networks and additionally have been shown to contribute to network stability: the three-species chain, omnivory, apparent competition, and direct competition \citep{Stouffer2007, Borrelli2015a, giling2019plant}.

@@

    \subsection*{Species persistence and motif participation}

       The relationship between frequency of participation in a motif and a species' probability of persistence depended on the level of disturbance. 
       When disturbance ($\pi$) was low, species' persistence increased with increasing proportions of the omnivory and direct competition motifs. 
       Conversely, persistence decreased with an increasing proportion of the apparent competition motif. 
       Persistence was not significantly related to the proportion of the three-species chain motif (Figure~\ref{fig:prop_lmer_all}; upper lines in each panel).

@@

    \subsection*{Motif participation and other network properties}

        Species' motif participation varied depending on the global properties of the network: size, connectance, and their interaction (Fig. S1, Table S2; \emph{Appendix S5}).
        Species appeared most often in the apparent competition motif in all types of networks, but the proportion of apparent competition decreased with increasing network size and connectance.
        The proportion of the omnivory motif in a species' participation role increased sharply with increasing connectance, while the proportion of the direct competition motif increased slightly with increasing network size.

        Species' motif participation also varied with their in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}).
        Species with higher in-degree tended to have higher frequencies of participation in the omnivory motif. Proportions of the other three motifs tended to decline with increasing degree.
        Species with higher trophic levels tended to have higher proportions of apparent competition and three-species chains. 
        A species' persistence varies with both species in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}; \emph{Appendix S6}). 
        Increased trophic level is generally associated with decreasing species persistence, although the rate of decline is slower at high disturbance (Table S3, \emph{Appendix S6}).
        In-degree shows an even stronger interaction with disturbance; having more prey species is associated with higher persistence without disturbance, but lower persistence at high levels of disturbance.

@@

            When fitting regressions of persistence against motif participation for each network separately, networks with high connectance generally showed more consistent relationships between persistence and motif participation, i.e., sharper peaks of the density distributions (Fig.~\ref{fig:density_prop_C}).
            However, this trend varied at different levels of disturbance, especially for the direct competition and omnivory motifs.
            At low and medium connectances there was little change in the distribution of slopes for direct competition across all levels of disturbance. 
            At high connectances, however, the proportion of positive slopes was both lower overall and decreased with increasing disturbance. 
            For the omnivory motif these trends were reversed. 
            Moreover, the omnivory motif has a particularly broad distribution of slopes (i.e., an inconsistent relationship to persistence) when network connectance was low. 
            Network size had a much smaller effect than connectance (Fig. S4; \emph{Appendix S7}).

    \subsection*{Network mean persistence and motif profiles}

        The overall motif profile of a network was related to the average persistence of consumers ($F_{1,2998}$=645, $p$\textless0.001), although this result must be considered tentative due to non-homogeneous distribution of roles (\emph{Appendix S8}). 
        Average persistence was higher in networks with higher proportions of the three-species chain, direct competition, and apparent competition motifs (Table S4, \emph{Appendix S8}). 
        Conversely, average persistence was lower in networks with higher proportions of the omnivory motif. 
        These trends were consistent across levels of disturbance (Fig.~\ref{fig:motif_profile_persistence}, Table S4), \emph{Appendix S8}).

@@
